\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2021-02-09 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
update manuscript

---
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James A. Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
abstract: |
  | In epidemiological studies of time-to-event data, a quantity of interest to the clinician and the patient is the risk of an event given their covariate profile. However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the likelihood expression or the estimating function. The baseline hazard then needs to be estimated separately using a non-parametric approach. This leads to step-wise estimates of the cumulative incidence that are difficult to interpret. Using case-base sampling, Hanley & Miettinen (2009) explained how the parametric hazard functions can be estimated using logistic regression. Their approach naturally leads to estimates of the cumulative incidence that are smooth-in-time.
  | * SRB and MT contributed equally to this work.
  | In this paper, we present the \CRANpkg{casebase} R package, a comprehensive and flexible toolkit for parametric survival analysis. We describe how the case-base framework can be used in more complex settings: non-linear functions of time and non-proportional hazards, competing risks, and variable selection. Our package also includes an extensive array of visualization tools to complement the analysis of time-to-event data. We illustrate all these features through three different case studies.
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{makecell}
  \usepackage{tabu}
  \DeclareUnicodeCharacter{2500}{-}
  \newcommand{\proglang}{\textsf}
output: rticles::rjournal_article 
bibliography: bhatnagar-turgeon-islam-hanley-saarela.bib 
fig_width: 12
fig_height: 9
fig_caption: true
editor_options: 
  chunk_output_type: console
---

<!-- # Code formatting -->

<!-- Don't use markdown, instead use the more precise latex commands: -->

<!-- * \proglang{Java} -->
<!-- * \pkg{plyr} -->
<!-- * \code{print("abc")} -->

@@

In this article, we present the \CRANpkg{casebase} package [@casebase-package] implemented in \proglang{R} [@r-core], for parametric survival analysis that combines the ideas of Hanley & Miettinen [-@hanley2009fitting] into a simple interface. The purpose of the \CRANpkg{casebase} package is to provide practitioners with an easy-to-use software tool to compute a patient's risk (or cumulative incidence) of an event, conditional on a particular patient's covariate profile. Our package retains the flexibility of case-base sampling and the familiar interface of the \code{glm} function. It also provides tools for variable selection, competing-risk analysis, and dealing with time-dependent exposures. In addition, we provide functions for exploratory data analysis and visualizing the estimated quantities.

In what follows, we first recall some theoretical details on case-base sampling and its use for estimating parametric hazard functions. We then give a short review of existing \proglang{R} packages that implement comparable features as \CRANpkg{casebase}. Next, we provide some details about the implementation of case-base sampling in our package, and we give a brief survey of its main functions. This is followed by three case studies that illustrate the flexibility and capabilities of \CRANpkg{casebase}. We show how the same framework can be used for non-linear functions of time and non-proportional hazards, competing risks and variable selection via penalized regression. Finally, we end the article with a discussion of the results and of future directions.

@@

To describe the theoretical foundations of case-base sampling, we use the framework of counting processes. In what follows, we abuse notation slightly and omit any mention of $\sigma$-algebras. Instead, following Aalen \emph{et al} [-@aalen2008survival], we use the placeholder "past" to denote the past history of the corresponding process. The reader interested in more details can refer to Saarela & Arjas [-@saarela2015non] and Saarela [-@saarela2016case]. First, let $N_{i}(t) \in \{0, 1\}$ be counting processes corresponding to the event of interest for individual $i=1, \ldots,n$. For simplicity, we will consider Type I censoring due to the end of follow-up at time $\tau$ (the general case of non-informative censoring is treated in Saarela [-@saarela2016case]). We assume a continuous time model, which implies that the counting process jumps are less than or equal to one. We are interested in modeling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
$$\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].$$

@@

In this model, the instantaneous failure rate is independent of $t$.\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study. This is also known as the \textit{memoryless property} \citep{kalbfleisch2011statistical}.}

@@

Survival analysis is an important branch of applied statistics and epidemiology. Accordingly, there is already a vast ecosystem of \proglang{R} packages implementing different methodologies. In this section, we describe how the functionalities of \CRANpkg{casebase} compare to these packages.

At the time of writing, a cursory examination of CRAN's \ctv{Survival} Task View reveals that there are over 250 packages related to survival analysis [@survTaskView]. For the purposes of this article, we restricted our review to packages that implement at least one of the following features: parametric modeling, non-proportional hazard models, competing risk analysis, penalized estimation, and CIF estimation. By searching for appropriate keywords in the \code{DESCRIPTION} file of these packages, we found 60 relevant packages. These 60 packages were then manually examined to determine which ones are comparable to \CRANpkg{casebase}. In particular, we excluded packages that were focused on a different set of problems, such as frailty and multi-state models. The remaining 14 packages appear in Table \ref{tab:surv-pkgs}, along with some of the functionalities they offer.

Parametric survival models are implemented in a handful of packages, each differing in the underlying data distribution: \CRANpkg{CFC} [-@mahani2015bayesian], \CRANpkg{flexsurv} [-@flexsurv], \CRANpkg{SmoothHazard} [-@smoothHazard], \CRANpkg{rsptm2} [-@clements_liu], \CRANpkg{mets} [-@scheike2014estimating], and \CRANpkg{survival} [-@survival-package]. For example, \CRANpkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas both \CRANpkg{flexsurv} and \CRANpkg{survival} allow users to supply any distribution of their choice. \CRANpkg{flexsurv}, \CRANpkg{smoothhazard}, \CRANpkg{mets} and \CRANpkg{rstpm2} can model the effect of time using splines, which allows flexible modeling of the hazard function. As discussed above, \CRANpkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear combination of covariates (including time). Therefore, our package is more general in that it allows the user to model any linear or non-linear transformation of time including splines and higher order polynomials. Also, by including interaction terms between covariates and time, it also allows users to fit (non-proportional) time-varying coefficient models. However, unlike \CRANpkg{flexsurv}, we do not explicitly model any shape parameter.

Several packages implement penalized estimation for the Cox model: \CRANpkg{glmnet} [-@regpathcox], \CRANpkg{glmpath} [-@park_hastie], \CRANpkg{penalized} [-@l1penal], \CRANpkg{riskRegression} [-@gerds_blanche]. Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: elastic-net penalization with \CRANpkg{rstpm2} [-@clements_liu], while \CRANpkg{survival} [-@survival-package] has an implementation of ridge-penalized estimation. On the other hand, our package \CRANpkg{casebase} provides penalized estimation of the hazard function. To our knowledge, \CRANpkg{casebase} and \CRANpkg{rsptm2} are the only packages to offer this functionality.

Next, several \proglang{R} packages implement methodologies for competing risk analysis; for a different perspective on this topic, see Mahani & Sharabiani [-@mahani2015bayesian]. The package \CRANpkg{cmprsk} provides methods for cause-specific subdistribution hazards, such as in the Fine-Gray model [-@fine1999proportional]. On the other hand, the package \CRANpkg{CFC} estimates cause-specific CIFs from unadjusted, non-parametric survival functions. Our package \CRANpkg{casebase} also provides functionalities for competing risk analysis by estimating parametrically the cause-specific hazards. From these quantities, we can then estimate the cause-specific CIFs.

Finally, several packages include functions to estimate the CIF. The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard. The first category broadly corresponds to parametric survival models, where the full hazard is explicitly modeled. Using this estimate, the survival function and the CIF can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below). Packages providing this functionality include \CRANpkg{CFC}, \CRANpkg{flexsurv}, \CRANpkg{mets}, and \CRANpkg{survival}. Our package \CRANpkg{casebase} also follows this approach for both single-event and competing-risk analyses. The second category outlined above broadly corresponds to semi-parametric models. These models do not model the full hazard function, and therefore the baseline hazard needs to be estimated separately in order to estimate the survival function. This is achieved using semi-parametric estimators (e.g. Breslow's estimator) or parametric estimators (e.g. spline functions). Packages that implement this approach include \CRANpkg{riskRegression}, \CRANpkg{rstpm2}, and \CRANpkg{survival}. As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the cumulative incidence function, whereas the second category often produces estimates in the form of stepwise functions. Providing smooth estimates of the CIF was one of the main motivations for introducing case-base sampling in survival analysis.

@@

\caption{Comparison of various \proglang{R} packages for survival analysis based on several defining features. \textbf{Competing Risks}: whether or not an implementation for competing risks is present. \textbf{Allows Non PH}: permits models for non-proportional hazards. \textbf{Penalized Regression}: allows for a penalty term on the regression coefficients when estimating hazards (e.g. lasso or ridge). \textbf{Splines}: permits a flexible fit on time through the use of splines. \textbf{Parametric}: implementation for parametric models. \textbf{Semi-parametric}: implementation for semi-parametric models. \textbf{Interval/left censoring}: models for interval and left-censoring. If this is not selected, the package only handles right-censoring. \textbf{Absolute Risk}: computation for survival curves, cumulative incidence or cumulative hazard is readily available.}

@@

The functions in the \CRANpkg{casebase} package can be divided into two categories: 1) exploratory data analysis, in the form of population-time plots; and 2) parametric modeling of the hazard function. We strove for compatibility with both \code{data.frame}s and \code{data.table}s; this can be seen in the coding choices we made and the unit tests we wrote.

@@

To illustrate these functions, we will use data from the European Randomized Study of Prostate Cancer Screening (ERSPC) [@schroder2009screening] which was extracted using the approach described in Liu *et al.* [-@liu2014recovering]. This dataset is available through the \CRANpkg{casebase} package. It contains the individual observations for 159,893 men  from seven European countries, who were between the ages of 55 and 69 years when recruited for the trial.

@@

  - \code{glm}: This is the familiar logistic regression.
  - \code{glmnet}: This option allows for variable selection using the elastic-net [@zou2005regularization] penalty (cf. Case Study 3). This functionality is provided through the \CRANpkg{glmnet} package [@friedman2010jss].
  - \code{gam}: This option provides support for *Generalized Additive Models* via the \CRANpkg{mgcv} package [@hastie1987generalized].
  - \code{gbm}: This option provides support for *Gradient Boosted Trees* via the \CRANpkg{gbm} package. This feature is still experimental.

@@

In this section, we illustrate the main functions of the \CRANpkg{casebase} package through three case studies. Each one showcases a different type of analysis. First, we show how to model different functional forms of time as well as non-proportional hazards. Then we perform a competing-risk analysis and compare our results with the Fine-Gray model. The third case study illustrates how to perform variable selection in high-dimensional datasets.

@@

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from \CRANpkg{casebase} on the Cox model CIF. The CIF estimates for the exponential model in panel (1) overestimate the cumulative incidence earlier on in time, and underestimate it later on. Based on our earlier discussion of the population-time plot, this poor fit for the exponential hazard was expected. We notice a better fit with increasing complexity of our model for time in Figure \ref{fig:erspc-cox-cif} (panels 2--4). As noted in Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models. Therefore, we can easily test the null hypothesis that the exponential model is just as good as the larger (in terms of number of parameters) spline model:

@@

<!--A common choice for this type of analysis is to use a Cox regression model and estimate the hazard ratio for the screening group (relative to the control group). In \proglang{R}, it can be done as follows:-->

@@

Previous re-analyses of these data suggested that the overall screening attributed reduction in death due to prostate cancer of 20% was an underestimate [@hanley2010mortality]. The estimated 20% (from a proportional hazards model) did not account for the delay between screening and the time the effect is expected to be observed. As a result, the null effects in years 1–7 masked the substantial reductions that began to appear from year 8 onward. This motivates the use of a time-dependent hazard ratio which can easily be fit with the \CRANpkg{casebase} package by including an interaction term with time in the model:

@@

In this case study, we show how case-base sampling can be used in the context of a competing-risk analysis. For illustrative purposes, we use the same data that was used in Scrucca *et al* [-@scrucca2010regression]. The data was downloaded from the first author's website, and it is now available as part of the \CRANpkg{casebase} package.

@@

For the third case study, we show how \CRANpkg{casebase} can also be used for variable selection through regularized estimation of the hazard function as given by \eqref{eq:penest}. To illustrate this functionality, we use the dataset from the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) [@knaus1995support]. The SUPPORT dataset tracks death in five American hospitals within individuals who are considered seriously ill. The original data is available online from the Department of Biostatistics at Vanderbilt University [@harrell_2020]. The cleaned and imputed data consists of 9104 observations and 30 variables, and it is available as part of the \CRANpkg{casebase} package. In the comparisons below, all covariates except \code{sps} and \code{aps} were used. These two variables correspond to scores for predicting the outcome that were developed as part of the original study. For more information about this dataset, the reader is encouraged to look at the documentation in our package.

For our penalized case-base model, we opt for the natural log of time which corresponds to a Weibull distribution. For fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface to the \code{fitSmoothHazard} function. We supply both a matrix \code{y} containing the time and event variables, and a matrix \code{x} containing all other covariates. We apply the lasso penalty by setting \code{alpha = 1} and assign a \code{penalty.factor} ($w_j$; cf. Equation \ref{eq:penest}) of 0 to the time variable to ensure it is in the selected model. We compare our approach to both Cox regression, and lasso penalized Cox regression (fitted via the \CRANpkg{glmnet} package).

To compare the performance of our models, we split the data into 95% training and 5% test sets. To assess both discrimination and calibration, we use a time-dependent version of the classical Brier score that is adjusted for censoring [@graf1999ass]. The Brier score can be used with both parametric and semi-parametric models. We use the \CRANpkg{riskRegression} package to compute these scores for all models.

@@

```{r coefplots, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3lolliPlot} Coefficient estimates from the Cox model (Cox), penalized Cox model using the \\CRANpkg{glmnet} package (Pen. Cox), and our approach using penalized case-base sampling (Pen. CB). Only the covariates that were selected by both penalized approaches are shown. The shrinkage of the coefficient estimates for Pen. Cox and Pen. CB occurs due to the $\\ell_1$ penalty."}
library(dotwhisker)
library(broom)
library(dplyr)

@@

In this article, we presented the \proglang{R} package \CRANpkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating CIFs using case-base sampling. Our package also provides several functions to produce graphical summaries of the data and the results. We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through three case studies.

@@

Even though we did not illustrate it in this article, case-base sampling can also be combined with the framework of *generalized additive models*. This functionality has already been implemented in the package. Similarly, case-base sampling can be combined with quasi-likelihood estimation to fit survival models that can account for the presence of over-dispersion. All of these examples illustrate how the case-base sampling framework in general, and the package \CRANpkg{casebase} in particular, allows the user to fit a broad and flexible family of survival functions.

@@

In the case studies above, we compared the performance of case-base sampling with that of Cox regression and Fine-Gray models. In terms of function interface, \CRANpkg{casebase} uses a formula interface that is closer to that of \code{glm}, in that the event variable is the only variable appearing on the left-hand side of the formula. By contrast, both \code{survival::coxph} and \code{timereg::comp.risk} use arrays that capture both the event type and time. Both approaches to modeling yield user-friendly code. However, in terms of output, both approaches differ significantly. Case-base sampling produces smooth hazards and smooth cumulative incidence curves, whereas Cox regression and Fine-Gray models produce stepwise CIFs and never explicitly model the hazard function. Qualitatively, we showed that by using splines in the linear predictor, all three models yielded similar curves. However, the smooth nature of the output of \CRANpkg{casebase} provides a more intuitive interpretation for consumers of these predictions. In Table \ref{tab:compCBvsCox}, we provide a side-by-side comparison between the Cox model and case-base sampling.

>>>>

---
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James A. Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
abstract: |
  | In epidemiological studies of time-to-event data, a quantity of interest to the clinician is their patient's risk of an event. However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the  estimating function. The baseline hazard then needs to be estimated separately using a non-parametric approach. This leads to step-wise estimates of the cumulative incidence that are difficult to interpret. Using case-base sampling, Hanley & Miettinen (2009) explained how parametric hazard functions can be estimated using logistic regression. Their approach naturally leads to estimates of the cumulative incidence that are smooth-in-time.
  | In this paper, we present the \CRANpkg{casebase} R package, a comprehensive and flexible toolkit for parametric survival analysis. We describe how the case-base framework can be used in more complex settings: non-linear functions of time and non-proportional hazards, competing risks, and variable selection. Our package also includes an extensive array of visualization tools to complement the analysis. We illustrate all these features through three different case studies.
  | * SRB and MT contributed equally to this work.
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{makecell}
  \usepackage{tabu}
  \DeclareUnicodeCharacter{2500}{-}
output: rticles::rjournal_article 
bibliography: bhatnagar-turgeon-islam-hanley-saarela.bib 
fig_width: 12
fig_height: 9
fig_caption: true
editor_options: 
  chunk_output_type: console
---

@@

In this article, we present the \CRANpkg{casebase} package [@casebase-package] implemented in R, for parametric survival analysis that combines the ideas of Hanley & Miettinen [-@hanley2009fitting] into a simple interface. The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to compute a patient's risk (or cumulative incidence) of an event, conditional on a particular patient's covariate profile. Our package retains the flexibility of case-base sampling and the familiar interface of the \code{glm} function. It also provides tools for variable selection, competing-risk analysis, and dealing with time-dependent exposures. In addition, we provide functions for exploratory data analysis and visualizing the estimated quantities.

In what follows, we first recall some theoretical details on case-base sampling and its use for estimating parametric hazard functions. We then give a short review of existing R packages that implement comparable features as \pkg{casebase}. Next, we provide some details about the implementation of case-base sampling in our package, and we give a brief survey of its main functions. This is followed by three case studies that illustrate the flexibility and capabilities of \pkg{casebase}. We show how the same framework can be used for non-linear functions of time and non-proportional hazards, competing risks and variable selection via penalized regression. Finally, we end the article with a discussion of the results and of future directions.

@@

To describe the theoretical foundations of case-base sampling, we use the framework of counting processes. In what follows, we abuse notation slightly and omit any mention of $\sigma$-algebras. Instead, following Aalen \textit{et al} [-@aalen2008survival], we use the placeholder "past" to denote the past history of the corresponding process. The reader interested in more details can refer to Saarela & Arjas [-@saarela2015non] and Saarela [-@saarela2016case]. First, let $N_{i}(t) \in \{0, 1\}$ be counting processes corresponding to the event of interest for individual $i=1, \ldots,n$. For simplicity, we will consider Type I censoring due to the end of follow-up at time $\tau$ (the general case of non-informative censoring is treated in Saarela [-@saarela2016case]). We assume a continuous time model, which implies that the counting process jumps are less than or equal to one. We are interested in modeling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
$$\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].$$

@@

In this model, the instantaneous failure rate is independent of $t$.\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study. This is also known as the \dfn{memoryless property} \citep{kalbfleisch2011statistical}.}

@@

Survival analysis is an important branch of applied statistics and epidemiology. Accordingly, there is already a vast ecosystem of R packages implementing different methodologies. In this section, we describe how the functionalities of \pkg{casebase} compare to these packages.

At the time of writing, a cursory examination of CRAN's \ctv{Survival} Task View reveals that there are over 250 packages related to survival analysis [@survTaskView]. For the purposes of this article, we restricted our review to packages that implement at least one of the following features: parametric modeling, non-proportional hazard models, competing risk analysis, penalized estimation, and CIF estimation. By searching for appropriate keywords in the \code{DESCRIPTION} file of these packages, we found 60 relevant packages. These 60 packages were then manually examined to determine which ones are comparable to \pkg{casebase}. In particular, we excluded packages that were focused on a different set of problems, such as frailty and multi-state models. The remaining 14 packages appear in Table \ref{tab:surv-pkgs}, along with some of the functionalities they offer.

Parametric survival models are implemented in a handful of packages, each differing in the underlying data distribution: \CRANpkg{CFC} [-@mahani2015bayesian], \CRANpkg{flexsurv} [-@flexsurv], \CRANpkg{SmoothHazard} [-@smoothHazard], \CRANpkg{rstpm2} [-@clements_liu], \CRANpkg{mets} [-@scheike2014estimating], and \CRANpkg{survival} [-@survival-package]. For example, \pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas both \pkg{flexsurv} and \pkg{survival} allow users to supply any distribution of their choice. \pkg{flexsurv}, \pkg{SmoothHazard}, \pkg{mets} and \pkg{rstpm2} can model the effect of time using splines, which allows flexible modeling of the hazard function. As discussed above, \pkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear combination of covariates (including time). Therefore, our package is more general in that it allows the user to model any linear or non-linear transformation of time including splines and higher order polynomials. Also, by including interaction terms between covariates and time, it also allows users to fit (non-proportional) time-varying coefficient models. However, unlike \pkg{flexsurv}, we do not explicitly model any shape parameter.

Several packages implement penalized estimation for the Cox model: \CRANpkg{glmnet} [-@regpathcox], \CRANpkg{glmpath} [-@park_hastie], \CRANpkg{penalized} [-@l1penal], \CRANpkg{riskRegression} [-@gerds_blanche]. Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: elastic-net penalization with \pkg{rstpm2} [-@clements_liu], while \pkg{survival} [-@survival-package] has an implementation of ridge-penalized estimation. On the other hand, our package \pkg{casebase} provides penalized estimation of the hazard function. To our knowledge, \pkg{casebase} and \pkg{rstpm2} are the only packages to offer this functionality.

Next, several R packages implement methodologies for competing risk analysis; for a different perspective on this topic, see Mahani & Sharabiani [-@mahani2015bayesian]. The package \CRANpkg{cmprsk} provides methods for cause-specific subdistribution hazards, such as in the Fine-Gray model [-@fine1999proportional]. On the other hand, the package \pkg{CFC} estimates cause-specific CIFs from unadjusted, non-parametric survival functions. Our package \pkg{casebase} also provides functionalities for competing risk analysis by estimating parametrically the cause-specific hazards. From these quantities, we can then estimate the cause-specific CIFs.

Finally, several packages include functions to estimate the CIF. The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard. The first category broadly corresponds to parametric survival models, where the full hazard is explicitly modeled. Using this estimate, the survival function and the CIF can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below). Packages providing this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows this approach for both single-event and competing-risk analyses. The second category outlined above broadly corresponds to semi-parametric models. These models do not model the full hazard function, and therefore the baseline hazard needs to be estimated separately in order to estimate the survival function. This is achieved using semi-parametric estimators (e.g. Breslow's estimator) or parametric estimators (e.g. spline functions). Packages that implement this approach include \pkg{riskRegression}, \pkg{rstpm2}, and \pkg{survival}. As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the cumulative incidence function, whereas the second category often produces estimates in the form of stepwise functions. Providing smooth estimates of the CIF was one of the main motivations for introducing case-base sampling in survival analysis.

@@

\caption{Comparison of various R packages for survival analysis based on several defining features. \textbf{Competing Risks}: whether or not an implementation for competing risks is present. \textbf{Allows Non PH}: permits models for non-proportional hazards. \textbf{Penalized Regression}: allows for a penalty term on the regression coefficients when estimating hazards (e.g. lasso or ridge). \textbf{Splines}: permits a flexible fit on time through the use of splines. \textbf{Parametric}: implementation for parametric models. \textbf{Semi-parametric}: implementation for semi-parametric models. \textbf{Interval/left censoring}: models for interval and left-censoring. If this is not selected, the package only handles right-censoring. \textbf{Absolute Risk}: computation for survival curves, cumulative incidence or cumulative hazard is readily available.}

@@

The functions in the \pkg{casebase} package can be divided into two categories: 1) exploratory data analysis, in the form of population-time plots; and 2) parametric modeling of the hazard function. We strove for compatibility with both \code{data.frame}s and \code{data.table}s; this can be seen in the coding choices we made and the unit tests we wrote.

@@

To illustrate these functions, we will use data from the European Randomized Study of Prostate Cancer Screening (ERSPC) [@schroder2009screening] which was extracted using the approach described in Liu *et al.* [-@liu2014recovering]. This dataset is available through the \pkg{casebase} package. It contains the individual observations for 159,893 men  from seven European countries, who were between the ages of 55 and 69 years when recruited for the trial.

@@

  - \code{glm}: This is the familiar logistic regression.
  - \code{glmnet}: This option allows for variable selection using the elastic-net [@zou2005regularization] penalty (cf. Case Study 3). This functionality is provided through the \pkg{glmnet} package [@friedman2010jss].
  - \code{gam}: This option provides support for *Generalized Additive Models* via the \CRANpkg{mgcv} package [@hastie1987generalized].
  - \code{gbm}: This option provides support for *Gradient Boosted Trees* via the \CRANpkg{gbm} package. This feature is still experimental.

@@

In this section, we illustrate the main functions of the \pkg{casebase} package through three case studies. Each one showcases a different type of analysis. First, we show how to model different functional forms of time as well as non-proportional hazards. Then we perform a competing-risk analysis and compare our results with the Fine-Gray model. The third case study illustrates how to perform variable selection in high-dimensional datasets.

@@

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from \pkg{casebase} on the Cox model CIF. The CIF estimates for the exponential model in panel (1) overestimate the cumulative incidence earlier on in time, and underestimate it later on. Based on our earlier discussion of the population-time plot, this poor fit for the exponential hazard was expected. We notice a better fit with increasing complexity of our model for time in Figure \ref{fig:erspc-cox-cif} (panels 2--4). As noted in Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models. Therefore, we can easily test the null hypothesis that the exponential model is just as good as the larger (in terms of number of parameters) spline model:

@@

<!--A common choice for this type of analysis is to use a Cox regression model and estimate the hazard ratio for the screening group (relative to the control group). In R, it can be done as follows:-->

@@

Previous re-analyses of these data suggested that the overall screening attributed reduction in death due to prostate cancer of 20% was an underestimate [@hanley2010mortality]. The estimated 20% (from a proportional hazards model) did not account for the delay between screening and the time the effect is expected to be observed. As a result, the null effects in years 1–7 masked the substantial reductions that began to appear from year 8 onward. This motivates the use of a time-dependent hazard ratio which can easily be fit with the \pkg{casebase} package by including an interaction term with time in the model:

@@

In this case study, we show how case-base sampling can be used in the context of a competing-risk analysis. For illustrative purposes, we use the same data that was used in Scrucca *et al* [-@scrucca2010regression]. The data was downloaded from the first author's website, and it is now available as part of the \pkg{casebase} package.

@@

For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function as given by \eqref{eq:penest}. To illustrate this functionality, we use the dataset from the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) [@knaus1995support].\footnote{The original data is available online from the Department of Biostatistics at Vanderbilt University: \url{https://biostat.app.vumc.org/wiki/Main/SupportDesc}} The SUPPORT dataset tracks death in five American hospitals within individuals who are considered seriously ill. The cleaned and imputed data consists of 9104 observations and 30 variables, and it is available as part of the \pkg{casebase} package. In the comparisons below, all covariates except \code{sps} and \code{aps} were used. These two variables correspond to scores for predicting the outcome that were developed as part of the original study. For more information about this dataset, the reader is encouraged to look at the documentation in our package.

For our penalized case-base model, we opt for the natural log of time which corresponds to a Weibull distribution. For fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface to the \code{fitSmoothHazard} function. We supply both a matrix \code{y} containing the time and event variables, and a matrix \code{x} containing all other covariates. We apply the lasso penalty by setting \code{alpha = 1} and assign a \code{penalty.factor} ($w_j$; cf. Equation \ref{eq:penest}) of 0 to the time variable to ensure it is in the selected model. We compare our approach to both Cox regression, and lasso penalized Cox regression (fitted via the \pkg{glmnet} package).

To compare the performance of our models, we split the data into 95% training and 5% test sets. To assess both discrimination and calibration, we use a time-dependent version of the classical Brier score that is adjusted for censoring [@graf1999ass]. The Brier score can be used with both parametric and semi-parametric models. We use the \pkg{riskRegression} package to compute these scores for all models.

@@

```{r coefplots, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3lolliPlot} Coefficient estimates from the Cox model (Cox), penalized Cox model using the \\pkg{glmnet} package (Pen. Cox), and our approach using penalized case-base sampling (Pen. CB). Only the covariates that were selected by both penalized approaches are shown. The shrinkage of the coefficient estimates for Pen. Cox and Pen. CB occurs due to the $\\ell_1$ penalty."}
library(dotwhisker)
library(broom)
library(dplyr)

@@

In this article, we presented the R package \pkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating CIFs using case-base sampling. Our package also provides several functions to produce graphical summaries of the data and the results. We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through three case studies.

@@

Even though we did not illustrate it in this article, case-base sampling can also be combined with the framework of *generalized additive models*. This functionality has already been implemented in the package. Similarly, case-base sampling can be combined with quasi-likelihood estimation to fit survival models that can account for the presence of over-dispersion. All of these examples illustrate how the case-base sampling framework in general, and the package \pkg{casebase} in particular, allows the user to fit a broad and flexible family of survival functions.

@@

In the case studies above, we compared the performance of case-base sampling with that of Cox regression and Fine-Gray models. In terms of function interface, \pkg{casebase} uses a formula interface that is closer to that of \code{glm}, in that the event variable is the only variable appearing on the left-hand side of the formula. By contrast, both \code{survival::coxph} and \code{timereg::comp.risk} use arrays that capture both the event type and time. Both approaches to modeling yield user-friendly code. However, in terms of output, both approaches differ significantly. Case-base sampling produces smooth hazards and smooth cumulative incidence curves, whereas Cox regression and Fine-Gray models produce stepwise CIFs and never explicitly model the hazard function. Qualitatively, we showed that by using splines in the linear predictor, all three models yielded similar curves. However, the smooth nature of the output of \pkg{casebase} provides a more intuitive interpretation for consumers of these predictions. In Table \ref{tab:compCBvsCox}, we provide a side-by-side comparison between the Cox model and case-base sampling.
